---
title: "Mapping America's Health: A County-Level Analysis Using CDC PLACES Data"
subtitle: "Unit 3 Group Project Report"
author: "Team 1 - Hazel Tran, Henry Phan"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    self-contained: true
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
bibliography: references.bib
---

## 1. Introduction

Chronic disease is the leading driver of death and disability in the United States. Conditions
like diabetes, obesity, and high blood pressure affect tens of millions of Americans, yet their
prevalence is not uniform. Where you live—your county—is one of the strongest predictors of
your health outcomes, shaped by the availability of preventive care, economic conditions,
and behavioral norms that vary dramatically from region to region.

This project asks three questions:

1. **Where** in the US are chronic disease rates highest, and are they geographically
   clustered?
2. **Does preventive care access** (measured as rates of annual checkups and cholesterol
   screening) associate with lower chronic disease prevalence at the county level?
3. **How strongly do health risk behaviors**—smoking and physical inactivity—predict
   chronic disease outcomes across counties?

To answer these questions, we use the **CDC PLACES: Local Data for Better Health** dataset
(2024 county-level release), which provides modeled estimates of 40 public health measures for
every one of the 3,143 counties and county-equivalents in the United States.

---

## 2. Data

### 2.1 Source

The CDC PLACES dataset is produced by the Centers for Disease Control and Prevention in
collaboration with the Robert Wood Johnson Foundation and the CDC Foundation. It is derived
from the Behavioral Risk Factor Surveillance System (BRFSS) and the US Census Bureau's
American Community Survey, using multilevel regression and post-stratification to generate
county-level estimates.

- **URL:** https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-County-Data-20/swc5-untb
- **Observations:** 3,143 rows (one per US county or county-equivalent)
- **Variables used:** 9 crude prevalence estimates + county identifiers + population
- **License:** US Government open data, public domain

### 2.2 Variables

| Variable | Description | Type |
|---|---|---|
| `diabetes_crudeprev` | % adults diagnosed with diabetes | Outcome |
| `obesity_crudeprev` | % adults with obesity (BMI ≥ 30) | Outcome |
| `highbp_crudeprev` | % adults with high blood pressure | Outcome |
| `csmoking_crudeprev` | % adults who currently smoke | Risk behavior |
| `lpa_crudeprev` | % adults with no leisure-time physical activity | Risk behavior |
| `checkup_crudeprev` | % adults with annual checkup in past year | Preventive care |
| `cholscreen_crudeprev` | % adults with cholesterol screening in past 5 yrs | Preventive care |
| `mhlth_crudeprev` | % adults reporting ≥14 poor mental health days/month | Health status |
| `phlth_crudeprev` | % adults reporting ≥14 poor physical health days/month | Health status |

### 2.3 Ethical Considerations

The PLACES dataset contains only aggregate, modeled estimates at the county level. No
individual-level records are present. The data is intentionally designed for public health
research and policy use, and all underlying survey data was collected with informed consent
through federal surveillance programs. We have no ethical concerns about using this dataset.

---

## 3. Exploratory Data Analysis

```{r}
#| label: setup-report
library(tidyverse)
library(scales)
library(corrplot)
library(janitor)

# Load data (long-format CDC API cache from EDA notebook)
places_long <- read_csv("../data/places_county.csv", show_col_types = FALSE) |>
  clean_names()

# Map measureid to column names for pivot
measure_id_map <- c(
  DIABETES   = "diabetes_crudeprev",
  OBESITY    = "obesity_crudeprev",
  BPHIGH     = "highbp_crudeprev",
  CHOLSCREEN = "cholscreen_crudeprev",
  CHECKUP    = "checkup_crudeprev",
  CSMOKING   = "csmoking_crudeprev",
  LPA        = "lpa_crudeprev",
  MHLTH      = "mhlth_crudeprev",
  PHLTH      = "phlth_crudeprev"
)

places <- places_long |>
  filter(measureid %in% names(measure_id_map),
         datavaluetypeid == "CrdPrv",
         !is.na(data_value)) |>
  mutate(data_value = as.numeric(data_value),
         col_name   = measure_id_map[measureid]) |>
  group_by(locationid) |>
  mutate(state_abbr       = first(stateabbr),
         state_desc       = first(statedesc),
         county_name      = first(locationname),
         total_population = as.numeric(first(totalpopulation))) |>
  ungroup() |>
  select(locationid, state_abbr, state_desc, county_name,
         total_population, col_name, data_value) |>
  pivot_wider(names_from = col_name, values_from = data_value,
              values_fn = mean) |>
  rename(county_fips = locationid) |>
  drop_na(state_abbr)
```

### 3.1 Regional Variation in Risk Behaviors

Smoking and physical inactivity — the two key risk behaviors in our dataset — vary
substantially across Census regions. The South shows consistently higher median rates
for both behaviors, while the West and Northeast tend to cluster lower. This regional
pattern foreshadows the geographic clustering of chronic disease outcomes explored in
Section 4.1.

```{r}
#| label: fig-region-boxplot
#| fig-cap: "Distribution of smoking and physical inactivity rates by Census region. The South shows the highest medians for both behaviors."

region_map <- c(
  CT="Northeast", ME="Northeast", MA="Northeast", NH="Northeast",
  RI="Northeast", VT="Northeast", NJ="Northeast", NY="Northeast",
  PA="Northeast",
  IL="Midwest",  IN="Midwest",  MI="Midwest",  OH="Midwest",
  WI="Midwest",  IA="Midwest",  KS="Midwest",  MN="Midwest",
  MO="Midwest",  NE="Midwest",  ND="Midwest",  SD="Midwest",
  DE="South",    FL="South",    GA="South",    MD="South",
  NC="South",    SC="South",    VA="South",    DC="South",
  WV="South",    AL="South",    KY="South",    MS="South",
  TN="South",    AR="South",    LA="South",    OK="South",
  TX="South",
  AZ="West",     CO="West",     ID="West",     MT="West",
  NV="West",     NM="West",     UT="West",     WY="West",
  AK="West",     CA="West",     HI="West",     OR="West",
  WA="West"
)

places |>
  mutate(region = region_map[state_abbr]) |>
  filter(!is.na(region)) |>
  select(region, csmoking_crudeprev, lpa_crudeprev) |>
  pivot_longer(-region, names_to = "measure", values_to = "prev") |>
  mutate(measure = recode(measure,
    csmoking_crudeprev = "Current Smoking",
    lpa_crudeprev      = "Physical Inactivity")) |>
  ggplot(aes(x = region, y = prev, fill = region)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.8, outlier.alpha = 0.3) +
  facet_wrap(~measure, scales = "free_y") +
  scale_fill_manual(values = c(
    Northeast = "#2980b9", South = "#c0392b",
    Midwest   = "#27ae60", West  = "#8e44ad"),
    guide = "none") +
  labs(x = NULL, y = "Crude Prevalence (%)") +
  theme_minimal(base_size = 11)
```

### 3.2 Geographic Variation

State-level averages reveal strong regional clustering. The highest diabetes and obesity rates
are concentrated in the Deep South (Mississippi, Alabama, Louisiana, West Virginia), while the
lowest are found in the Mountain West and Northeast. This "Diabetes Belt" pattern has been
documented in public health literature and is visible clearly in the data.

```{r}
#| label: fig-state-diabetes
#| fig-cap: "State average county diabetes prevalence. Southern states consistently lead."

places |>
  group_by(state_abbr) |>
  summarise(avg_diabetes = mean(diabetes_crudeprev, na.rm = TRUE)) |>
  ggplot(aes(x = reorder(state_abbr, avg_diabetes), y = avg_diabetes,
             fill = avg_diabetes)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "#74add1", high = "#d73027", guide = "none") +
  labs(x = NULL, y = "Average Diabetes Prevalence (%)") +
  theme_minimal(base_size = 10)
```

---

## 4. Research Findings

### 4.1 Question 1: Geographic Clustering

Yes, chronic disease rates are strongly clustered geographically. The five states with the
highest average county diabetes prevalence are all in the South or Appalachian region. This
pattern holds for obesity and high blood pressure as well. The geographic clustering suggests
that shared environmental, economic, and cultural factors — not individual behavior alone —
drive these outcomes.

### 4.2 Question 2: Prevention and Outcomes

Counties with higher rates of annual checkups and cholesterol screening show a modest *positive*
correlation with chronic disease prevalence. This counterintuitive finding — more prevention
associated with more disease — likely reflects a detection bias: counties with better healthcare
access diagnose and record more cases. It also reflects that preventive care tends to be
deployed more in populations with known risk. This is a classic confounding problem in
observational health data and highlights why causal claims cannot be made from this dataset alone.

```{r}
#| label: fig-prevention-scatter
#| fig-cap: "Checkup rate vs. diabetes prevalence. Higher checkup rates correlate with higher detected diabetes — likely a detection effect."

places |>
  ggplot(aes(x = checkup_crudeprev, y = diabetes_crudeprev)) +
  geom_point(alpha = 0.25, color = "#2c7bb6", size = 1) +
  geom_smooth(method = "lm", color = "#d73027", se = TRUE) +
  labs(x = "Annual Checkup Rate (%)", y = "Diabetes Prevalence (%)") +
  theme_minimal()
```

### 4.3 Question 3: Risk Behaviors and Outcomes

Physical inactivity and current smoking both show strong positive associations with diabetes
and obesity prevalence across counties. The correlation between physical inactivity and
diabetes is particularly strong (r ≈ 0.75), suggesting that counties where adults report
low physical activity are also counties with high diabetes rates. Again, this is a correlation
and not a causal claim — both may be driven by shared socioeconomic factors.

```{r}
#| label: fig-inactivity
#| fig-cap: "Physical inactivity vs. diabetes prevalence at the county level."

places |>
  ggplot(aes(x = lpa_crudeprev, y = diabetes_crudeprev)) +
  geom_point(alpha = 0.25, color = "#4dac26", size = 1) +
  geom_smooth(method = "lm", color = "#d73027", se = TRUE) +
  labs(x = "Physical Inactivity Rate (%)", y = "Diabetes Prevalence (%)") +
  theme_minimal()
```

---

## 5. Dashboard Design

### 5.1 Design Decisions

Our Shiny dashboard is a **single-page layout** organized around the three research questions.
Every chart is visible at once, with global controls (state filter, measure selector) at the
top. The six charts are:

- **Chart 1 — Choropleth Map** (Q1): A state-level choropleth showing average prevalence of
  the selected measure, colored on a diverging blue-to-red scale. This gives the user an
  immediate geographic orientation and reveals regional clustering.
- **Chart 2 — Top 10 Bar** (Q1): A horizontal bar chart of the 10 highest-prevalence counties
  for precise ranking — bar length is easier to compare than map shading.
- **Chart 3 — Scatter + Trend Line** (Q2): An interactive scatter plot pairing any preventive
  care measure against any outcome, with an optional linear trend line. Users can switch axes
  to explore checkup rate vs. diabetes, cholesterol screening vs. obesity, etc.
- **Chart 4 — Boxplot by Care Access** (Q2): Counties are split at the median care rate into
  "High Access" and "Low Access" groups. Side-by-side boxplots show how outcome distributions
  differ between groups.
- **Chart 5 — Small Multiples by Region** (Q3): Faceted bar charts showing average smoking
  and physical inactivity rates across the four Census regions (Northeast, Midwest, South,
  West), enabling quick regional comparison of risk behaviors.
- **Chart 6 — Lollipop Chart** (Q3): For a user-selected state and county, shows each
  county's deviation from the state average — red stems above the line, blue below. This
  lets users benchmark individual counties against their peers.

### 5.2 Visualization Principles

Following the Unit 2 principles, we made these deliberate choices:

- **Perceptual ranking:** We use bar charts for comparison tasks (rankings, state comparisons)
  because position on a common scale is the most accurately perceived encoding.
- **Color:** We use a single diverging palette (blue → red) to encode magnitude. We avoid
  rainbow palettes. All charts use `theme_minimal()` to reduce chartjunk.
- **Accessibility:** Color is never the only encoding — bars also encode position and length.
  The scatter plot uses transparency (alpha) to handle overplotting.
- **Interactivity with purpose:** Every interactive element (state filter, measure selector,
  scatter axis selector) changes what question is being answered, not just the appearance.

---

## 6. Limitations

1. **Ecological fallacy.** All variables are county-level aggregates. We cannot infer
   individual-level relationships from this data.
2. **Modeled estimates.** PLACES values are model-based, not direct survey counts. Counties
   with small populations have wider uncertainty intervals.
3. **Confounding.** We have not controlled for income, rurality, or demographic composition,
   all of which are likely confounders.
4. **Cross-sectional data.** We cannot establish temporal precedence or causation.

---

## 7. Conclusion

The CDC PLACES data reveals stark geographic inequality in chronic disease burden across US
counties. Diabetes, obesity, and high blood pressure are heavily concentrated in the South
and Appalachia, and are strongly associated with physical inactivity and smoking at the
county level. Our dashboard makes these patterns accessible and explorable for a non-technical
audience, allowing users to investigate any state or measure of interest. The most important
takeaway for public health is that these are not randomly distributed problems — they are
concentrated in specific places, which means targeted geographic interventions are both
possible and necessary.

---

## References

Centers for Disease Control and Prevention. (2024). *PLACES: Local Data for Better Health,
County Data 2024 Release.* https://data.cdc.gov/500-Cities-Places/
PLACES-Local-Data-for-Better-Health-County-Data-20/swc5-untb
