---
title: "CDC PLACES — Exploratory Data Analysis"
subtitle: "CP2: Data Loading & EDA"
author: "Your Team Name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

## 1. Setup & Data Loading

```{r}
#| label: setup

library(tidyverse)   # data wrangling & ggplot2
library(janitor)     # clean column names
library(skimr)       # quick summary stats
library(corrplot)    # correlation matrix
library(scales)      # axis formatting

# ---------------------------------------------------------
# Download the CDC PLACES county-level CSV directly.
# The Socrata open data API lets us pull CSV with a URL.
# If this fails, download manually from:
# https://data.cdc.gov/500-Cities-Places/
#   PLACES-Local-Data-for-Better-Health-County-Data-20/swc5-untb
# and place it in the data/ folder as "places_county.csv"
# ---------------------------------------------------------

data_url <- paste0(
  "https://data.cdc.gov/resource/swc5-untb.csv",
  "?$limit=4000"   # all counties (3143 rows), add buffer
)

places_raw <- read_csv(data_url) |>
  clean_names()   # snake_case all column names

# Save locally so you don't re-download each time
write_csv(places_raw, "data/places_county.csv")

cat("Rows:", nrow(places_raw), "\n")
cat("Columns:", ncol(places_raw), "\n")
```

## 2. Data Shape & Quality

```{r}
#| label: skim

# Quick overview of every column
skim(places_raw)
```

```{r}
#| label: missing-values

# Visualise missingness
places_raw |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  arrange(desc(pct_missing)) |>
  ggplot(aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col(fill = "#2c7bb6") +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Missing Values by Column",
       x = NULL, y = "% Missing") +
  theme_minimal()
```

## 3. Key Variables

We focus on the **crude prevalence** estimates (columns ending in `_crudeprev`) which express
the estimated percentage of adults in each county with a given condition.

```{r}
#| label: select-vars

# Select the measures we care about most
places <- places_raw |>
  select(
    state_abbr,
    state_desc,
    county_name,
    county_fips,
    total_population,
    # Chronic disease outcomes
    diabetes_crudeprev,
    obesity_crudeprev,
    highbp_crudeprev,
    # Preventive care
    cholscreen_crudeprev,
    colon_screen_crudeprev,
    checkup_crudeprev,
    # Risk behaviors
    csmoking_crudeprev,
    lpa_crudeprev,          # physical inactivity
    binge_crudeprev,
    # Health status
    mhlth_crudeprev,
    phlth_crudeprev
  ) |>
  # Convert all prevalence columns to numeric
  mutate(across(ends_with("_crudeprev"), as.numeric),
         total_population = as.numeric(total_population))

glimpse(places)
```

## 4. Distributions

### 4a. Chronic Disease Outcomes

```{r}
#| label: outcome-distributions

places |>
  select(diabetes_crudeprev, obesity_crudeprev, highbp_crudeprev) |>
  pivot_longer(everything(), names_to = "measure", values_to = "prevalence") |>
  mutate(measure = recode(measure,
    diabetes_crudeprev = "Diabetes",
    obesity_crudeprev  = "Obesity",
    highbp_crudeprev   = "High Blood Pressure"
  )) |>
  ggplot(aes(x = prevalence, fill = measure)) +
  geom_histogram(bins = 40, alpha = 0.85, color = "white") +
  facet_wrap(~measure, scales = "free_x") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(title = "Distribution of Chronic Disease Prevalence Across US Counties",
       x = "Crude Prevalence (%)", y = "Number of Counties") +
  theme_minimal()
```

### 4b. Risk Behaviors

```{r}
#| label: behavior-distributions

places |>
  select(csmoking_crudeprev, lpa_crudeprev, binge_crudeprev) |>
  pivot_longer(everything(), names_to = "measure", values_to = "prevalence") |>
  mutate(measure = recode(measure,
    csmoking_crudeprev = "Current Smoking",
    lpa_crudeprev      = "Physical Inactivity",
    binge_crudeprev    = "Binge Drinking"
  )) |>
  ggplot(aes(x = prevalence, fill = measure)) +
  geom_histogram(bins = 40, alpha = 0.85, color = "white") +
  facet_wrap(~measure, scales = "free_x") +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  labs(title = "Distribution of Health Risk Behaviors Across US Counties",
       x = "Crude Prevalence (%)", y = "Number of Counties") +
  theme_minimal()
```

## 5. Top & Bottom Counties

```{r}
#| label: top-bottom-diabetes

bind_rows(
  places |> slice_max(diabetes_crudeprev, n = 10) |> mutate(group = "Highest"),
  places |> slice_min(diabetes_crudeprev, n = 10) |> mutate(group = "Lowest")
) |>
  mutate(label = paste0(county_name, ", ", state_abbr)) |>
  ggplot(aes(x = reorder(label, diabetes_crudeprev),
             y = diabetes_crudeprev, fill = group)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Highest" = "#d73027", "Lowest" = "#4575b4")) +
  labs(title = "Counties with Highest and Lowest Diabetes Prevalence",
       x = NULL, y = "Diabetes Prevalence (%)", fill = NULL) +
  theme_minimal()
```

## 6. Correlations

```{r}
#| label: correlations

cor_data <- places |>
  select(diabetes_crudeprev, obesity_crudeprev, highbp_crudeprev,
         csmoking_crudeprev, lpa_crudeprev,
         cholscreen_crudeprev, checkup_crudeprev) |>
  drop_na()

cor_matrix <- cor(cor_data)

corrplot(cor_matrix,
         method   = "color",
         type     = "upper",
         tl.cex   = 0.8,
         addCoef.col = "black",
         number.cex  = 0.7,
         col = colorRampPalette(c("#4575b4", "white", "#d73027"))(200),
         title = "Correlation Matrix: Health Outcomes & Behaviors",
         mar = c(0,0,2,0))
```

## 7. State-Level Summary

```{r}
#| label: state-summary

places |>
  group_by(state_abbr) |>
  summarise(
    n_counties        = n(),
    avg_diabetes      = mean(diabetes_crudeprev, na.rm = TRUE),
    avg_obesity       = mean(obesity_crudeprev,  na.rm = TRUE),
    avg_smoking       = mean(csmoking_crudeprev, na.rm = TRUE),
    avg_inactivity    = mean(lpa_crudeprev,       na.rm = TRUE)
  ) |>
  arrange(desc(avg_diabetes)) |>
  slice_head(n = 15) |>
  knitr::kable(digits = 1,
    caption = "Top 15 States by Average County Diabetes Prevalence")
```

## 8. Research Questions

Based on the EDA above, we will answer the following questions in our final dashboard:

1. **Geographic distribution:** Which US counties and states have the highest rates of
   diabetes, obesity, and high blood pressure? Are these clustered regionally?

2. **Prevention vs. outcomes:** Do counties with higher rates of preventive care (annual
   checkups, cholesterol screening) show lower chronic disease prevalence?

3. **Behavior vs. outcomes:** How strongly do smoking and physical inactivity predict
   diabetes and obesity at the county level?
